A Simplified Spectral Deferred Correction Method for Coupling Hydrodynamics with Reaction Networks and Nuclear Statistical Equilibrium

### 2020-10-16 simplified_sdc_intro/paper.tex
some updates

\newcommand{\nablab}{{\mathbf{\nabla}}}
\newcommand{\Ub}{\mathbf{U}}
\newcommand{\gb}{\mathbf{g}}
\newcommand{\omegadot}{\dot{\omega}}
\newcommand{\Sdot}{\dot{S}}
\newcommand{\ddx}[1]{{\frac{{\partial#1}}{\partial x}}}
\newcommand{\ddt}[1]{{\frac{{\partial#1}}{\partial t}}}
\newcommand{\odt}[1]{{\frac{{d#1}}{dt}}}
\newcommand{\divg}[1]{{\nablab \cdot \left (#1\right)}}
\newcommand{\inp}{{\mathrm{input}}}
\newcommand{\dedr}{\left . {\partial{}e}/{\partial\rho}\right |_{T, X_k}}
\newcommand{\dedrd}{\left . \frac{\partial{}e}{\partial\rho}\right |_{T, X_k}}
\newcommand{\dedX}{\left . {\partial{}e}/{\partial{}X_k} \right |_{\rho, T}}
\newcommand{\dedXd}{\left . \frac{\partial{}e}{\partial{}X_k} \right |_{\rho, T, X_{j,j\ne k}}}
\newcommand{\dedT}{\left . {\partial{}e}/{\partial{}T} \right |_{\rho,X_k}}
\newcommand{\dedTd}{\left . \frac{\partial{}e}{\partial{}T} \right |_{\rho,X_k}}

@@

@@

The NSE table provides:
\begin{align}
Y_e &= \sum_k \frac{Z_k X_k}{A_k} \\
\bar{A} &= \left [ \sum_k \frac{X_k}{A_k} \right ]^{-1} \\
\frac{B}{A} &= \sum_k \frac{B_k X_k}{A_k}
\end{align}
where $B_k$ is the binding energy of nucleus $k$.  In our simulations,
we store these 3 quantities as auxiliary data that is carried along
with the rest of the fluid state in each zone.

and our evolution equations are:
\begin{align}
\frac{DY_e}{Dt} &= \sum_k \frac{Z_k}{A_k} \frac{DX_k}{Dt} = \sum_k \frac{Z_k}{A_k} \omegadot_k \\
\frac{D\bar{A}}{Dt} &= -\bar{A}^2 \sum_k \frac{1}{A_k} \frac{DX_k}{Dt} = -\bar{A}^2 \sum_k \frac{1}{A_k} \omegadot_k \\
\frac{D}{Dt} \left (\frac{B}{A} \right ) &= \sum_k \frac{B_k}{A_k} \frac{DX_k}{Dt} = \sum_k \frac{B_k}{A_k} \omegadot_k
\end{align}
For Strang split coupling of hydro and reactions, $DX_k/Dt = 0$,
Therefore each of these auxillary equations obeys an advection
equation in the hydro part of the advancement.  In the SDC algorithm,
there will be a reactive source (an $\Ic_q$) for each of these that is
computed in the same manner as above.

@@

For Strang splitting, the basic flow of using the aprox19 + NSE network
is as follows:
\begin{itemize}
\item {\em Initialize the problem}

  We initializing the mass fractions, $X_k$, and then compute the
  electron fraction, $Y_e$, from them.  For the two other composition
  quantities we carry, $\bar{A}$, and $(B/A)$, we need values that are
  consistent with the value of $Y_e$ and the nuclei stored in the NSE
  table.  Therefore, if the thermodynamic conditions put us in NSE
  (using the conditions defined below), then we get $\bar{A}$ and
  $(B/A)$ from the NSE table.  Otherwise, we compute these directly
  from $X_k$.  $X_k$.

@@

  \begin{itemize}
  \item For a zone that is in NSE:

    \begin{itemize}
    \item use $\rho$, $T$, and $Y_e$ to call the table. This
      returns: $dY_e/dt$, $(B/A)_{\rm out}$, $\bar{A}_{\rm out}$,
      and the $(X_k)_{\rm out}$ from the NSE network grouped into the 19 isotopes
      we carry as part of aprox19.

    \item update $Y_e$:
    \begin{equation}
      (Y_e)_{\rm out} = (Y_e)_{\rm in} + \Delta t \frac{dY_e}{dt}
    \end{equation}

  \item update the auxiliary state value of $\bar{A}$ to be simply
    $\bar{A}_{\rm out}$ from the table and the mass fractions
    $X_k$ to be the $(X_k)_{\rm out}$ returned from the table.

  \item compute the energy generation rate, $\Sdot$ as:
    \begin{equation}
      \Sdot = \eta \left [ \left ( \frac{B}{A} \right )_{\rm out} -
        \left ( \frac{B}{A} \right )_{\rm in} \right ] N_A \frac{1}{\Delta t}
    \end{equation}
    where $\eta$ is an inertia term < 1 to prevent the energy changing
    too much in one set.

  \item set the auxiliary state value of $(B/A)$ as:
    \begin{equation}
      \left (\frac{B}{A}\right )_{\rm out} = \left (\frac{B}{A}\right )_{\rm in} + \eta \left [ \left ( \frac{B}{A}\right )_{\rm out} - \left (\frac{B}{A} \right )_{\rm in} \right]
    \end{equation}

@@

@@

\acknowledgements \castro\ is freely available at
\url{http://github.com/AMReX-Astro/Castro}.  All of the code and
problem setups used here are available in the git repo.  The work at
Stony Brook was supported by DOE/Office of Nuclear Physics grant
DE-FG02-87ER40317.  This material is based upon work supported by the
U.S. Department of Energy, Office of Science, Office of Advanced
Scientific Computing Research and Office of Nuclear Physics, Scientific
Discovery through Advanced Computing (SciDAC) program under Award
Number DE-SC0017955.

>>>>

\newcommand{\nablab}{{\mathbf{\nabla}}}
\newcommand{\Ub}{\mathbf{U}}
\newcommand{\gb}{\mathbf{g}}
\newcommand{\omegadot}{\dot{\omega}}
\newcommand{\Sdot}{\dot{S}}
\newcommand{\ddx}[1]{{\frac{{\partial#1}}{\partial x}}}
\newcommand{\ddt}[1]{{\frac{{\partial#1}}{\partial t}}}
\newcommand{\odt}[1]{{\frac{{d#1}}{dt}}}
\newcommand{\divg}[1]{{\nablab \cdot \left (#1\right)}}
\newcommand{\dedr}{\left . {\partial{}e}/{\partial\rho}\right |_{T, X_k}}
\newcommand{\dedrd}{\left . \frac{\partial{}e}{\partial\rho}\right |_{T, X_k}}
\newcommand{\dedX}{\left . {\partial{}e}/{\partial{}X_k} \right |_{\rho, T}}
\newcommand{\dedXd}{\left . \frac{\partial{}e}{\partial{}X_k} \right |_{\rho, T, X_{j,j\ne k}}}
\newcommand{\dedT}{\left . {\partial{}e}/{\partial{}T} \right |_{\rho,X_k}}
\newcommand{\dedTd}{\left . \frac{\partial{}e}{\partial{}T} \right |_{\rho,X_k}}

@@

\newcommand{\nse}[1]{{\mathrm{NSE}( #1 )}}

\newcommand{\out}{{\rm out}}
\newcommand{\inp}{{\rm in}}

@@

The NSE table requires:
\begin{equation}
  Y_e = \sum_k \frac{Z_k X_k}{A_k}
\end{equation}
(where $A_k$ and $Z_k$ are the atomic weight and atomic number of nucleus $k$) and provides
\begin{align}
\bar{A} &= \left [ \sum_k \frac{X_k}{A_k} \right ]^{-1} \\
\left (\frac{B}{A} \right ) &= \sum_k \frac{B_k X_k}{A_k}
\end{align}
where $B_k$ is the binding energy of nucleus $k$.  In our simulations,
we store these 3 quantities as auxiliary data that is carried along
with the rest of the fluid state in each zone.  The table also returns
values of the mass fractions mapped onto the 19-isotopes we carry, $X_k$, and the time-derivative of $Y_e$, $dY_e/dt$.
We'll use the notation:
\begin{equation}
\nse{\rho,T,Y_e} \rightarrow \bar{A}, X_k, (B/A), dY_e/dt
\end{equation}
to represent the NSE table call and its inputs and outputs.

We can derive evolution equations for each of these composition quantities as:
\begin{align}
\frac{DY_e}{Dt} &= \sum_k \frac{Z_k}{A_k} \frac{DX_k}{Dt} = \sum_k \frac{Z_k}{A_k} \omegadot_k \\
\frac{D\bar{A}}{Dt} &= -\bar{A}^2 \sum_k \frac{1}{A_k} \frac{DX_k}{Dt} = -\bar{A}^2 \sum_k \frac{1}{A_k} \omegadot_k \\
\frac{D}{Dt} \left (\frac{B}{A} \right ) &= \sum_k \frac{B_k}{A_k} \frac{DX_k}{Dt} = \sum_k \frac{B_k}{A_k} \omegadot_k
\end{align}
For Strang split coupling of hydro and reactions, $DX_k/Dt = 0$,
Therefore each of these auxillary equations obeys an advection
equation in the hydro part of the advancement.  In the SDC algorithm,
there will be a reactive source (an $\Ic_q$) for each of these that is
computed in the same manner as above.  We note that our NSE table
provides the evolution of $Y_e$ due to reactions ($DY_e/Dt$) directly.

@@

\subsection{Initialization}

We initializing the mass fractions, $X_k$, and then compute the
electron fraction, $Y_e$, from them.  For the two other composition
quantities we carry, $\bar{A}$, and $(B/A)$, we need values that are
consistent with the value of $Y_e$ and the nuclei stored in the NSE
table.  Therefore, if the thermodynamic conditions put us in NSE
(using the conditions defined below), then we get $\bar{A}$ and
$(B/A)$ from the NSE table.  Otherwise, we compute these directly from
$X_k$.

\subsection{NSE condition}

\subsection{Strang-split algorithm for NSE}

For Strang splitting, the basic flow of using the aprox19 + NSE network
is as follows:
\begin{itemize}

@@

    \item For a zone that is in NSE:

      The goal is to update the composition and thermodynamics due to the 
      change in the nuclei abundances over the (half-)timestep.  There are
      a few approaches we can do here.

      \begin{itemize}
      \item Method I: {\em Use the time-level $n$ information for the update}.

        This is the approach used in \citet{ma:2013}.  We call the table with the 
        input state:
        \begin{equation}
          \nse{\rho, T, (Y_e)_{\rm in}} \rightarrow \bar{A}_\out, (X_k)_\out, (B/A)_\out, dY_e/dt
        \end{equation}

        We can now update $Y_e$:
        \begin{equation}
          (Y_e)_{\rm out} = (Y_e)_{\rm in} + \Delta t \frac{dY_e}{dt}
        \end{equation}
        and take the auxiliary state values of $\bar{A}$ and $(B/A)$ to be simply
        the values from the table, $\bar{A}_\out$ and $(B/A)_\out$, and the mass fractions
        $X_k$ to be the $(X_k)_{\rm out}$ returned from the table.

        Finally, we compute the energy generation rate, $\Sdot$ as:
        \begin{equation}
          \label{eq:nse_energy}
          \Sdot = \left [ \left ( \frac{B}{A} \right )_{\rm out} -
            \left ( \frac{B}{A} \right )_{\rm in} \right ] N_A \frac{1}{\Delta t}
        \end{equation}

        We note that \citet{ma:2013} modified Eq.~\ref{eq:nse_energy}
        to take use only a fraction (0.3) of the energy change from
        the table, to avoid a potential instability that can arise if
        too much energy is deposited in a zone in a single timestep
        (they called this an inertia term).  We prefer to deal with
        the issue of too much energy being deposited in a zone in one
        step through the retry mechanism already built into \castro.

      \item Method II: {\em predictor-corrector.}

        This starts out the same as the above method:
        \begin{equation}
          \nse{\rho, T, (Y_e)_{\rm in}} \rightarrow \bar{A}^\star, (X_k)^\star, (B/A)^\star, dY_e/dt
        \end{equation}
        but we indicate with a $\star$ that most of those values are
        provisional, and we will seek a better value in the corrector step.
        We only update $Y_e$ from this call:
        \begin{equation}
          (Y_e)_{\rm out} = (Y_e)_{\rm in} + \Delta t \frac{dY_e}{dt}
        \end{equation}
        and then we use the table again, but with the updated $Y_e$:
        \begin{equation}
          \nse{\rho, T, (Y_e)_\out} \rightarrow \bar{A}_\out, (X_k)_\out, (B/A)_\out, dY_e/dt
        \end{equation}
        This corrects the composition so that it is consistent with the updated
        $Y_e$.  We use these values $\bar{A}_\out$, $(X_k)_\out$, and $(B/A)_\out$ to then complete the update, computing the energy as in Eq.~\ref{eq:nse_energy}.

@@

\subsection{SDC-NSE Coupling}

@@

\acknowledgements \castro\ is freely available at
\url{http://github.com/AMReX-Astro/Castro}.  All of the code and
problem setups used here are available in the git repo.  The work at
Stony Brook was supported by DOE/Office of Nuclear Physics grant
DE-FG02-87ER40317.  This material is based upon work supported by the
U.S. Department of Energy, Office of Science, Office of Advanced
Scientific Computing Research and Office of Nuclear Physics,
Scientific Discovery through Advanced Computing (SciDAC) program under
Award Number DE-SC0017955.  This research was supported by the
Exascale Computing Project (17-SC-20-SC), a collaborative effort of
the U.S. Department of Energy Office of Science and the National
Nuclear Security Administration.
